---
title: "Education Under Fire: Boko Haram Conflict and Educational Attainment in Northern Nigeria"
author: "Jarret"
date: today
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 12pt
    linestretch: 1.5
    number-sections: true
    toc: true
    toc-depth: 3
    colorlinks: true
  docx:
    toc: true
    number-sections: true
bibliography: references.bib
csl: chicago-author-date.csl
execute:
  echo: false        # hide code in output
  warning: false
  message: false
  cache: true        # cache R output so re-renders are fast
---

```{r setup}
#| include: false

# Source the shared config — loads all packages and sets paths
source(here::here("R", "00_config.R"))

# Load the analysis panel
panel <- readr::read_csv(PATH_PANEL, show_col_types = FALSE) |>
  dplyr::mutate(
    lga_id      = factor(lga_id),
    survey_year = as.integer(survey_year)
  )

# Load pre-estimated model objects (produced by 03_main_did.R)
models <- readRDS(file.path(PATH_TABLES, "main_models.rds"))
```

# Introduction

Nigeria's northeast has experienced one of the world's most destructive
insurgencies since the Boko Haram movement began its violent campaign in
the late 2000s. The conflict has displaced millions of people, destroyed
infrastructure, and disrupted educational systems at a critical juncture
for the region's development [@unocha2020].

This paper estimates the causal effect of Boko Haram conflict exposure on
educational attainment in Northern Nigeria. Using a difference-in-differences
(DiD) design, I compare educational outcomes in Local Government Areas (LGAs)
with high pre-conflict Boko Haram exposure to those with low exposure, before
and after the insurgency intensified in 2009.

**Research question:** Did exposure to Boko Haram conflict causally reduce
school enrolment, attendance, and years of completed schooling in
Northern Nigeria?

**Main finding:** [Update after estimation] Conflict exposure reduced years
of schooling by approximately X years and school enrolment rates by
approximately Y percentage points, with effects concentrated among girls
and children who were school-age in 2009.

## Contribution

This paper makes three contributions to the literature on conflict and
human capital:

1. **Causal identification**: Unlike much of the descriptive literature, I
   exploit the sharp geographic variation in Boko Haram's pre-insurgency
   presence to implement a credible DiD design.

2. **Disaggregated outcomes**: I separately estimate effects on enrolment,
   attendance, and completed years of schooling, allowing me to distinguish
   between the intensive and extensive margins of educational disruption.

3. **Heterogeneity**: I show that girls, Muslim-majority LGAs, and children
   who were school-age in 2009 experienced systematically larger effects.

---

# Background and Context

## Boko Haram's Insurgency

Boko Haram ("Western education is forbidden") emerged as a religious movement
in Maiduguri, Borno State in the early 2000s under Mohammed Yusuf.
The movement's violent phase began following clashes with security forces
in 2009, the year used as the treatment onset in this paper
[@thurston2017boko].

The insurgency's geographic reach has been concentrated in the Lake Chad
Basin region — particularly Borno, Yobe, and Adamawa states — but
spillover effects have been documented across the broader northeast and
into the northwest.

## Education in Northern Nigeria

Pre-conflict educational attainment in Northern Nigeria was already
substantially below southern levels, driven by a combination of poverty,
cultural factors, and limited school supply [@combined2019]. The conflict
compounded these pre-existing deficits through:

- Direct school destruction and school closures
- Teacher displacement and absenteeism
- Household income shocks reducing the ability to pay school fees
- Direct targeting of students and teachers by Boko Haram

---

# Data

## ACLED Conflict Event Data

Conflict event data come from the Armed Conflict Location and Event Data
project (ACLED), which records geo-coded conflict events from news reports
and other sources [@raleigh2010]. I use events classified as:
battles, explosions/remote violence, and violence against civilians, for
Nigeria, 2000–2020.

## DHS Survey Microdata

Educational outcome data come from the Demographic and Health Surveys (DHS)
programme. I use the Individual Recode files from Nigeria's DHS rounds
conducted between `r min(panel$survey_year, na.rm=TRUE)` and
`r max(panel$survey_year, na.rm=TRUE)`.

Key variables extracted from the DHS:

- **Years of schooling**: total years of formal education completed
- **Current enrolment**: indicator for currently attending school
- **Primary completion**: indicator for having completed primary school
- **GPS cluster coordinates**: used to assign respondents to LGAs

## Summary Statistics

```{r summary-stats}
#| label: tbl-summary
#| tbl-cap: "Descriptive Statistics"

modelsummary::datasummary(
  (`Years of schooling`  = years_of_schooling)  +
  (`Enrolled`            = enrolled)             +
  (`Primary complete`    = primary_complete)     +
  (`IHS fatalities`      = ihs_fatalities)       +
  (`Conflict events`     = n_events)             ~
    N + Mean + SD + Min + Max,
  data  = panel,
  fmt   = 2,
  notes = "Sample: Northern Nigeria. DHS microdata merged with ACLED conflict data."
)
```

The analysis covers `r n_distinct(panel$lga_id)` unique LGAs and
`r format(nrow(panel), big.mark=",")` individual respondents.
Treated LGAs (those with more than `r EXPOSURE_THRESH` pre-2009 conflict
events) account for
`r scales::percent(mean(panel$treated, na.rm=TRUE), accuracy=1)` of
the sample.

---

# Identification Strategy

## Difference-in-Differences Design

The main estimating equation is:

$$
Y_{ilt} = \alpha + \beta \cdot (\text{Treated}_l \times \text{Post2009}_t) + \gamma_l + \delta_t + X_{it}'\theta + \varepsilon_{ilt}
$$

where $Y_{ilt}$ is the educational outcome for individual $i$ in LGA $l$
at survey year $t$; $\text{Treated}_l$ equals one if LGA $l$ had more than
`r EXPOSURE_THRESH` conflict events before 2009; $\text{Post2009}_t$
equals one for survey years 2009 and later; $\gamma_l$ are LGA fixed
effects; $\delta_t$ are survey-year fixed effects; and $X_{it}$ is a
vector of individual controls (age, age², urban/rural).

The coefficient of interest, $\beta$, is the average treatment effect on
the treated (ATT): the additional change in educational outcomes experienced
by treated LGAs after 2009, relative to what the control group experienced.

Standard errors are clustered at the LGA level to account for serial
correlation of outcomes within LGAs over time.

## Parallel Trends Assumption

The DiD design requires that, absent the Boko Haram insurgency, treated
and control LGAs would have followed parallel trends. This assumption is
untestable for the post-period but can be evaluated for the pre-period.

@fig-pretrends shows the mean educational trends by treatment group from
`r YEAR_START` to `r TREATMENT_YEAR - 1`. [Describe what you observe.]

```{r pretrend-plot}
#| label: fig-pretrends
#| fig-cap: "Pre-Trend Check: Educational Outcomes by Treatment Group"
#| fig-height: 5

knitr::include_graphics(file.path(PATH_FIGURES, "fig_pretrends.pdf"))
```

---

# Results

## Main DiD Estimates

@tbl-main presents the main DiD estimates. Columns (1), (3), and (5) show
the baseline TWFE specification without individual controls; columns (2),
(4), and (6) add age, age², and urban/rural controls.

```{r main-table}
#| label: tbl-main
#| tbl-cap: "Main DiD Results: Effect of Conflict Exposure on Education"

model_list <- list(
  "(1)" = models[["years_of_schooling"]]$spec1,
  "(2)" = models[["years_of_schooling"]]$spec2,
  "(3)" = models[["enrolled"]]$spec1,
  "(4)" = models[["enrolled"]]$spec2,
  "(5)" = models[["primary_complete"]]$spec1,
  "(6)" = models[["primary_complete"]]$spec2
)

modelsummary::modelsummary(
  model_list,
  coef_map = c("treated_post" = "Treated × Post-2009",
               "age"          = "Age",
               "I(age^2)"     = "Age²"),
  stars    = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  gof_map  = c("nobs", "adj.r.squared"),
  notes    = "SEs clustered by LGA. *** p<0.01, ** p<0.05, * p<0.10."
)
```

## Event Study

@fig-event-study presents the event study estimates, showing the dynamic
pattern of treatment effects in each year relative to 2009.

```{r event-study}
#| label: fig-event-study
#| fig-cap: "Event Study: Dynamic Treatment Effects"
#| fig-height: 8

knitr::include_graphics(file.path(PATH_FIGURES, "fig_event_study.pdf"))
```

Pre-period coefficients are [close to zero / statistically insignificant],
consistent with the parallel trends assumption. Post-period coefficients
show [describe pattern].

## Heterogeneity

```{r heterogeneity}
#| label: fig-het
#| fig-cap: "Heterogeneity Analysis: DiD Estimates by Subgroup"
#| fig-height: 5

knitr::include_graphics(file.path(PATH_FIGURES, "fig_heterogeneity.pdf"))
```

---

# Robustness Checks

@tbl-robust shows that the main results are robust to:

1. **Alternative conflict thresholds** — the DiD estimate is stable
   across thresholds of 2, 5, 10, and 20 pre-period events.
2. **Continuous conflict measures** — replacing the binary treatment
   indicator with IHS-transformed fatality counts yields similar results.
3. **Placebo treatment years** — assigning fake treatment years of 2003,
   2005, and 2007 produces coefficients near zero, as expected.

```{r robustness-table}
#| label: tbl-robust
#| tbl-cap: "Robustness Checks"

knitr::include_graphics(file.path(PATH_FIGURES, "fig_threshold_sensitivity.pdf"))
```

---

# Conclusion

[To be completed after estimation.]

This paper provides causal evidence that Boko Haram conflict exposure
reduced educational attainment in Northern Nigeria. The results have
implications for [policy implications] and suggest that [main takeaway].

---

# References {.unnumbered}

::: {#refs}
:::
